---
title: "Replication: Violence Trends in the Ancient Middle East"
author: "Replication Assistant"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

## 1. Introduction & Setup

This document replicates the analysis from Baten, Benati & Sołtysiak (2023), *Violence trends in the ancient Middle East between 12,000 and 400 BCE*.

We adhere to strict computational reproducibility by translating the original Stata code (`0redo_viol230707nhb.do`) 1:1 into R.

### Load Libraries

We use standard `library()` calls to be compatible with `renv` or standard R environments.

```{r libraries}
library(tidyverse)  # Data manipulation and plotting
library(censReg)    # Tobit regression (Maximum Likelihood estimation, matches Stata)
library(sandwich)   # Robust Standard Errors
library(lmtest)     # Coefficient testing

```

## 2. Data Loading and Preparation

We load the dataset `S1_file_combined.xlsx - Sheet1.csv`. Note that this file contains aggregated site-level data () with pre-generated dummy variables for periods and regions.

```{r data_prep}
# 1. Load Data
# Ensure the filename matches exactly what is in your directory
df <- readxl::read_xlsx("S1_file_combined.xlsx")

head(df)
```

```{r}

df$period <- factor(df$LF, ordered = FALSE)

report::report_table(df)

vars <- c('rcc', 'royal', 'urban', 'period', 'battle', 'iron', 'horse', 'size', 'trauma', 'tr_nop')

```


```{r}

df_model <- df %>%
  clean_names() %>% # Standardizes names (e.g., T0n_ -> t0n, T1c_ -> t1c)
  
  # --- Reconstruct Period Factor ---
  # The CSV provides dummies: t0n (Neolithic), t1c (Chalcolithic), etc.
  # We convert these back into a single factor variable for the regression.
  mutate(
    period_raw = case_when(
      t0n == 1 ~ "Neolithic",
      t1c == 1 ~ "Chalcolithic",
      t2be == 1 ~ "Early Bronze",
      t3bm == 1 ~ "Middle Bronze",
      t4bl == 1 ~ "Late Bronze",
      t5i == 1  ~ "Iron Age",
      TRUE ~ NA_character_
    ),
    # Set Factor Levels explicitly so "Chalcolithic" is the reference (intercept)
    period_factor = factor(period_raw, levels = c(
      "Chalcolithic",      # REFERENCE CATEGORY
      "Neolithic", 
      "Early Bronze", 
      "Middle Bronze", 
      "Late Bronze", 
      "Iron Age"
    ))
  ) %>%
  
  # --- Reconstruct Region Factor ---
  # The 'co2' column in the CSV contains region codes:
  # l_ = Levant, ms = Mesopotamia, tr = Turkey, ir = Iran
  mutate(
    region = case_when(
      co2 == "l_" ~ "Levant",
      co2 == "ms" ~ "Mesopotamia",
      co2 == "tr" ~ "Turkey",
      co2 == "ir" ~ "Iran",
      TRUE ~ "Unknown"
    ),
    region = factor(region)
  ) %>%
  
  # --- Create Dependent & Control Variables ---
  mutate(
    # Dependent Variable: log(violence_share + 0.05)
    # The paper explicitly adds 0.05 to handle zeros before logging.
    # 'trauma' in the CSV represents the raw violence share.
    lviol = log(trauma + 0.05),
    
    # Control Variables (Ensure numeric 0/1)
    urban_dummy = as.numeric(urban),
    battle_dummy = as.numeric(battle),
    royal_dummy = as.numeric(royal),
    
    # Tech variables (Iron, Horse)
    iron_dummy = as.numeric(iron),
    horse_dummy = as.numeric(horse),
    
    # Climate Shock (Variable 'rcc' = Rapid Climate Change/Drought share)
    climate_shock = rcc,
    
    # Population Size (Variable 'size' in CSV)
    population = size
  )

head(df_model)

```


```{r}
# Define the censoring limit used in Stata code (log of the added constant)
cens_limit <- log(0.05)

# --- ESTIMATE MODEL 3 ---
# Specification: Period Dummies + Urban + Battle + Royal + Region Fixed Effects
m1 <- lviol ~ period_factor

# Run Tobit (Maximum Likelihood)
m1_tobit <- censReg(m1, left = cens_limit, data = df_model)

# --- CALCULATE ROBUST SEs ---
# Replicating Stata's "vce(cluster region)"
# We use vcovCL to get the clustered covariance matrix
m1_vcov <- vcovCL(m1_tobit)
```

```{r}
summary(m1_tobit)
summary(m1_vcov)

report::report_table(m1_tobit)
```

