---
title: "Replication: Violence Trends in the Ancient Middle East"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

## 1. Introduction & Setup

This document replicates the analysis from Baten, Benati & Sołtysiak (2023), *Violence trends in the ancient Middle East between 12,000 and 400 BCE*.

The goal is to translate the original Stata code (`0redo_viol230707nhb.do`) into R while maintaining exact computational reproducibility for the variable transformations, model specifications (Tobit), and standard errors.

### Load Libraries

```{r libraries}

library(tidyverse)
library(janitor)
library(censReg)
library(sandwich)
library(lmtest)
library(knitr)
library(kableExtra)

```

## 2. Data Loading and Preparation

We load the dataset and perform the exact data cleaning steps found in the Stata do-file.

**Stata Logic Replicated:**

1. 
**Filtering:** Keep years between 12,000 BCE and 400 BCE.


2. 
**Transformation:** The paper calculates violence intensity as `log(violence_share + 0.05)` to handle zeros in the Tobit model.


3. 
**Factor Levels:** The "Chalcolithic" period is set as the reference category.



```{r data_prep}
# Load the raw data
# Ensure the CSV file name matches your local file
d_raw <- read_csv("S1_file_combined.xlsx - Sheet1.csv")

# Clean and Transform
d_model <- d_raw %>%
  clean_names() %>% # Converts names to snake_case (e.g., "Violence Share" -> "violence_share")
  
  # Filter for the relevant period (12,000 - 400 BCE)
  filter(year_bce >= 400 & year_bce <= 12000) %>%
  
  # --- EXACT REPLICATION OF VARIABLES ---
  
  # 1. Dependent Variable: Log transformation with constant 0.05
  # Stata: gen lviol = log(viol + 0.05)
  mutate(lviol = log(violence_share + 0.05)) %>%
  
  # 2. Period Factors (Reference = Chalcolithic)
  # We manually set levels to ensure Chalcolithic is the baseline intercept.
  mutate(
    period_factor = factor(period, levels = c(
      "Chalcolithic",      # Reference Category
      "Neolithic", 
      "Early Bronze", 
      "Middle Bronze", 
      "Late Bronze", 
      "Iron Age"
    ))
  ) %>%
  
  # 3. Control Variable Conversions
  # Ensure dummies are numeric (0/1) for regression input
  mutate(
    urban_dummy = as.numeric(urban),
    battle_dummy = as.numeric(battle),
    royal_dummy = as.numeric(royal),
    
    # Handle missing values for tech variables if they exist in raw data
    # (Stata treats missing as missing, but for regression checks we assume 0 if NA implies absence)
    iron_dummy = ifelse(!is.na(iron), as.numeric(iron), 0),
    horse_dummy = ifelse(!is.na(horse), as.numeric(horse), 0),
    
    # Climate Shock: Defined as share of period with drought
    climate_shock = if("drought_share" %in% names(.)) drought_share else NA
  )

# Verify the structure
glimpse(d_model)

```

## 3. Statistical Modeling (Tobit)

We replicate **Table 2, Model 3** from the paper. This is the primary model discussed in the text, controlling for Region, Battle sites, Royal burials, and Urbanism.

* **Method:** Tobit Regression (Censored Regression).
* **Censoring Point:** Since the DV is `log(y + 0.05)`, the left-censoring limit is `log(0.05)`.
* **Standard Errors:** Clustered by Region (Stata: `vce(cluster region)`).

```{r analysis_tobit}
# Define the censoring limit (log(0.05) ≈ -2.9957)
cens_limit <- log(0.05)

# --- MODEL SPECIFICATION ---
# Formula: lviol ~ Period + Urban + Battle + Royal + Region
# Note: "Region" is added as a factor to create fixed effects.
f_m3 <- lviol ~ period_factor + urban_dummy + battle_dummy + royal_dummy + factor(region)

# --- ESTIMATION (Maximum Likelihood) ---
# This matches Stata's 'tobit' command
m3_tobit <- censReg(f_m3, left = cens_limit, data = d_model)

# --- ROBUST STANDARD ERRORS ---
# Replicating Stata's 'vce(cluster region)'
# We use vcovCL from the sandwich package
m3_vcov <- vcovCL(m3_tobit, cluster = d_model$region)

```

### Regression Results

Below is the replication of the coefficients and significance levels. Note that `factor(region)` coefficients are often suppressed in paper tables but are estimated in the background.

```{r results_table}
# Generate the coefficient test table using the clustered covariance matrix
m3_results <- coeftest(m3_tobit, vcov = m3_vcov)

# Display as a tidy table
m3_results %>%
  tidy() %>%
  mutate(
    significance = case_when(
      p.value < 0.01 ~ "***",
      p.value < 0.05 ~ "**",
      p.value < 0.1 ~ "*",
      TRUE ~ ""
    )
  ) %>%
  kable(caption = "Replication of Table 2, Model 3 (Tobit with Clustered SEs)") %>%
  kable_styling(full_width = F)

```

**Interpretation Check:**
If the coefficient for `period_factorNeolithic` is approx **-1.332** and `period_factorMiddle Bronze` is approx **-1.815**, the replication is successful.

## 4. Visualization (Figure 2)

We replicate the "Spaghetti Plot" showing violence trends by region and the overall Middle East average.

```{r plot_fig2, fig.width=10, fig.height=6}
# 1. Aggregate Data for Plotting
plot_data <- d_model %>%
  group_by(region, period_factor) %>%
  summarise(
    mean_viol = mean(violence_share, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# 2. Calculate the "Middle East" Grand Mean
grand_mean <- d_model %>%
  group_by(period_factor) %>%
  summarise(
    mean_viol = mean(violence_share, na.rm = TRUE),
    region = "Middle East (Avg)"
  )

# 3. Combine for plotting
# We add the grand mean to the main data for a unified legend
plot_combined <- bind_rows(plot_data, grand_mean)

# 4. Create the Plot
ggplot(plot_combined, aes(x = period_factor, y = mean_viol, color = region, group = region)) +
  # Lines for regions
  geom_line(aes(linetype = region == "Middle East (Avg)", size = region == "Middle East (Avg)")) +
  geom_point() +
  
  # Customizing styles to match the paper's aesthetic
  scale_size_manual(values = c(1, 1.5), guide = "none") + # Thicker line for Average
  scale_linetype_manual(values = c("solid", "dashed"), guide = "none") +
  
  # Labels
  labs(
    title = "Trends in Interpersonal Violence (12,000 - 400 BCE)",
    subtitle = "Replication of Figure 2: Mean violence share by region and period",
    y = "Violence Share (%)",
    x = "Period",
    color = "Region"
  ) +
  
  # Theme adjustments
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

```

## 5. Sensitivity Checks (Preview)

While the above replicates the paper's exact method, the use of `log(y + 0.05)` on proportion data is methodologically questionable.

A more robust approach (to be implemented in future steps) would be:

1. **Binomial GLM:** Modeling the counts of injuries directly using `glm(cbind(injured, total-injured) ~ ..., family = binomial)`.
2. **Zero-Inflated Models:** Checking if the zeros (0% violence) are structural.

```

```