---
title: "Tobit (intreg-style) vs Beta-binomial: Diagnostics & Model Comparison"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}

options(scipen = 999)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

suppressPackageStartupMessages({
  library(readxl)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

have_pkg <- function(p) requireNamespace(p, quietly = TRUE)

round_df <- function(x, digits = 3) {
  x %>% mutate(across(where(is.numeric), ~ round(.x, digits)))
}

add_goodfit_subtitle <- function(txt) {
  mtext(txt, side = 3, line = 0.2, cex = 0.85)
}

good_line_txt <- "Good fit: points close to 45° line; no systematic over/under-prediction across periods."
```

## 1) Data and shared preprocessing

```{r data}
# Update path if needed
# df_raw <- read_excel("S1_file_combined.xlsx")
df_raw <- read_excel("S1_file_combined.xlsx")

# Analysis dataset (Model 3 covariates)
dat <- df_raw %>%
  transmute(
    tr_nop = as.numeric(tr_nop),
    ncases = as.numeric(ncases),
    co2LF  = as.factor(co2LF),
    co2    = as.factor(co2),
    T0n_   = as.numeric(`T0n_`),
    T2be   = as.numeric(T2be),
    T3bm   = as.numeric(T3bm),
    T4bl   = as.numeric(T4bl),
    T5i_   = as.numeric(`T5i_`),
    battle = as.numeric(battle),
    royal  = as.numeric(royal),
    urban  = as.numeric(urban)
  ) %>%
  filter(
    !is.na(tr_nop), !is.na(ncases), !is.na(co2LF), !is.na(co2),
    !is.na(T0n_), !is.na(T2be), !is.na(T3bm), !is.na(T4bl), !is.na(T5i_),
    !is.na(battle), !is.na(royal), !is.na(urban),
    ncases > 0
  )

stopifnot(nrow(dat) == 82)

# Match Stata base category if relevant
if ("ir" %in% levels(dat$co2)) dat$co2 <- relevel(dat$co2, ref = "ir")

# Period labels (used later)
dat <- dat %>%
  mutate(
    period = case_when(
      T0n_ == 1 ~ "T0n_",
      T2be == 1 ~ "T2be",
      T3bm == 1 ~ "T3bm",
      T4bl == 1 ~ "T4bl",
      T5i_ == 1 ~ "T5i_",
      TRUE ~ "BASE"
    )
  )

# Shared design matrix for Tobit/intreg model
X <- model.matrix(
  ~ T0n_ + T2be + T3bm + T4bl + T5i_ + battle + royal + urban + co2,
  data = dat
)
K <- ncol(X)
N <- nrow(dat)

cluster <- dat$co2LF
```

## 2) Tobit / intreg-style model (Stata-replication likelihood)

### 2.1 Fit model

```{r tobit-fit}
# Tobit/intreg outcome construction (unchanged)
y <- ifelse(dat$tr_nop > 0, log(dat$tr_nop), NA_real_)
is_cens <- dat$tr_nop == 0

# Stata-normalized analytic weights: sum(a) = N (unchanged)
a <- dat$ncases
stopifnot(all(is.finite(a)), all(a > 0))
a <- a * N / sum(a)

# Left-censor point (baseline)
cpoint <- -3

loglik <- function(par) {
  beta <- par[1:K]
  lns  <- par[K + 1]
  s <- exp(lns)
  xb <- as.vector(X %*% beta)

  ll <- numeric(N)

  # Uncensored
  idx_u <- which(!is_cens)
  r <- y[idx_u] - xb[idx_u]
  ll[idx_u] <- -0.5 * ( a[idx_u] * (r^2) / (s^2) + log(2*pi) + 2*lns - log(a[idx_u]) )

  # Left-censored
  idx_c <- which(is_cens)
  t <- (cpoint - xb[idx_c]) * sqrt(a[idx_c]) / s
  ll[idx_c] <- pnorm(t, log.p = TRUE)

  sum(ll)
}

start <- c(rep(0, K), 0)
fit <- optim(start, fn = function(p) -loglik(p), method = "BFGS", hessian = TRUE)
stopifnot(fit$convergence == 0)

theta <- fit$par
beta_hat <- theta[1:K]
lnsigma_hat <- theta[K + 1]
sigma_hat <- exp(lnsigma_hat)

xb <- as.vector(X %*% beta_hat)

# Per-observation scores U for cluster-robust sandwich
U <- matrix(0, nrow = N, ncol = K + 1)
colnames(U) <- c(colnames(X), "lnsigma")

# Uncensored scores
idx_u <- which(!is_cens)
r <- y[idx_u] - xb[idx_u]
U[idx_u, 1:K]   <- (a[idx_u] * r / sigma_hat^2) * X[idx_u, , drop = FALSE]
U[idx_u, K + 1] <- -1 + (a[idx_u] * r^2 / sigma_hat^2)

# Censored scores
idx_c <- which(is_cens)
tc <- (cpoint - xb[idx_c]) * sqrt(a[idx_c]) / sigma_hat
log_phi <- dnorm(tc, log = TRUE)
log_Phi <- pnorm(tc, log.p = TRUE)
lambda <- exp(pmin(log_phi - log_Phi, 700))

U[idx_c, 1:K]   <- -(lambda * sqrt(a[idx_c]) / sigma_hat) * X[idx_c, , drop = FALSE]
U[idx_c, K + 1] <- -(lambda * tc)

# Cluster-robust sandwich
bread <- solve(fit$hessian)
Sg <- rowsum(U, cluster)
meat <- crossprod(Sg)
V_CR0 <- bread %*% meat %*% bread

G <- nrow(Sg)
V_stata <- (G/(G - 1)) * V_CR0  # Stata-style small-sample correction

Vb <- V_stata[1:K, 1:K, drop = FALSE]
se <- sqrt(diag(Vb))
zv <- beta_hat / se
pv <- 2 * pnorm(-abs(zv))

out_tobit <- data.frame(
  term = colnames(X),
  estimate = beta_hat,
  se_cr = se,
  z = zv,
  p = pv
)

cat("Tobit/intreg baseline fit complete.\n")
round_df(out_tobit, 3)
```

### 2.2 Tobit diagnostics (fit, censoring calibration, surprises)

```{r tobit-diagnostics}
ll_hat <- loglik(theta)
k_par <- K + 1
AIC_t <- -2 * ll_hat + 2 * k_par
BIC_t <- -2 * ll_hat + log(N) * k_par

# Wald chi^2 for slopes (robust)
idx_slope <- which(colnames(X) != "(Intercept)")
b_slope <- beta_hat[idx_slope]
V_slope <- Vb[idx_slope, idx_slope, drop = FALSE]
Wald_chi2 <- as.numeric(t(b_slope) %*% solve(V_slope) %*% b_slope)
df_wald <- length(idx_slope)
p_wald <- pchisq(Wald_chi2, df = df_wald, lower.tail = FALSE)

# Model-implied P(censored)
sd_i <- sigma_hat / sqrt(a)
p_cens_hat <- pnorm((cpoint - xb) / sd_i)

# Observed vs predicted censoring rates
obs_cens_unw  <- mean(is_cens)
pred_cens_unw <- mean(p_cens_hat)
obs_cens_w    <- weighted.mean(is_cens, w = a)
pred_cens_w   <- weighted.mean(p_cens_hat, w = a)

# Brier score for censoring indicator
brier_w <- weighted.mean((as.numeric(is_cens) - p_cens_hat)^2, w = a)

# RMSE among uncensored (descriptive)
res_u <- y[!is_cens] - xb[!is_cens]
rmse_u <- sqrt(weighted.mean(res_u^2, w = a[!is_cens]))

# Null model for McFadden pseudo-R2
loglik0 <- function(par0) {
  b0 <- par0[1]
  lns0 <- par0[2]
  s0 <- exp(lns0)
  xb0 <- rep(b0, N)

  ll0 <- numeric(N)
  idx_u0 <- which(!is_cens)
  r0 <- y[idx_u0] - xb0[idx_u0]
  ll0[idx_u0] <- -0.5 * ( a[idx_u0] * (r0^2) / (s0^2) + log(2*pi) + 2*lns0 - log(a[idx_u0]) )

  idx_c0 <- which(is_cens)
  t0 <- (cpoint - xb0[idx_c0]) * sqrt(a[idx_c0]) / s0
  ll0[idx_c0] <- pnorm(t0, log.p = TRUE)

  sum(ll0)
}

fit0 <- optim(c(0, 0), fn = function(p) -loglik0(p), method = "BFGS", hessian = TRUE)
stopifnot(fit0$convergence == 0)
ll_null <- loglik0(fit0$par)
pseudoR2 <- 1 - (ll_hat / ll_null)

# Surprise ranking (large values = surprising given model)
p_unc_hat <- 1 - p_cens_hat
surprise <- ifelse(is_cens, -log(p_cens_hat), -log(p_unc_hat))
w_surprise <- a * surprise
ord <- order(w_surprise, decreasing = TRUE)

fit_stats_tobit <- data.frame(
  N = N,
  uncensored = sum(!is_cens),
  left_censored = sum(is_cens),
  clusters = G,
  logLik = ll_hat,
  AIC = AIC_t,
  BIC = BIC_t,
  Wald_chi2 = Wald_chi2,
  df = df_wald,
  p_value = p_wald,
  obs_cens_unw = obs_cens_unw,
  pred_cens_unw = pred_cens_unw,
  obs_cens_w = obs_cens_w,
  pred_cens_w = pred_cens_w,
  brier_w = brier_w,
  rmse_uncensored = rmse_u,
  sigma = sigma_hat,
  ll_null = ll_null,
  pseudoR2_mcfadden = pseudoR2,
  max_w_surprise = max(w_surprise)
)

round_df(fit_stats_tobit, 3)

top_surprise <- data.frame(
  row = ord[1:10],
  weight_a = a[ord[1:10]],
  xb = xb[ord[1:10]],
  is_cens = is_cens[ord[1:10]],
  p_cens = p_cens_hat[ord[1:10]],
  w_surprise = w_surprise[ord[1:10]]
)

round_df(top_surprise, 3)
```

### 2.3 Tobit diagnostic plots

```{r tobit-plots}
op <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))

plot(
  xb[!is_cens], res_u,
  xlab = "Fitted xb (latent mean)",
  ylab = "Residual (y - xb) [uncensored]",
  main = sprintf("Tobit/intreg (cpoint=%s): Residuals vs fitted", cpoint)
)
abline(h = 0, lty = 2)
add_goodfit_subtitle("Good fit: residuals centered at 0; no strong pattern/heteroskedasticity across xb.")

boxplot(
  p_cens_hat ~ is_cens,
  names = c("Uncensored", "Censored"),
  ylab = "Predicted P(censored)",
  main = sprintf("Tobit/intreg (cpoint=%s): P(censored) by observed status", cpoint)
)
add_goodfit_subtitle("Good fit: censored cases have higher predicted P(censored) than uncensored cases.")

par(op)
```

## 3) Beta-binomial model (glmmTMB) on reconstructed counts

### 3.1 Reconstruct integer counts

```{r reconstruct-counts}
dat2 <- dat %>%
  select(period, everything()) %>%   # <-- ensure period is present
  mutate(
    n = as.integer(round(ncases)),
    p = tr_nop,
    k_raw = p * n,
    k = as.integer(round(k_raw)),
    k = pmin(pmax(k, 0L), n)
  )

stopifnot("period" %in% names(dat2))
stopifnot(all(dat2$n > 0), all(dat2$p >= 0 & dat2$p <= 1))


recon_check <- dat2 %>%
  mutate(diff = k_raw - k) %>%
  summarise(
    max_abs_diff = max(abs(diff)),
    mean_abs_diff = mean(abs(diff)),
    share_exact_integer = mean(abs(diff) < 1e-8)
  )

round_df(recon_check, 3)
```

### 3.2 Fit beta-binomial and run diagnostics

```{r betabin-fit}
if (!have_pkg("glmmTMB")) stop("Please install glmmTMB")
library(glmmTMB)

f_bin <- cbind(k, n - k) ~ T0n_ + T2be + T3bm + T4bl + T5i_ + battle + royal + urban + co2

m_bb <- glmmTMB(
  f_bin,
  data = dat2,
  family = betabinomial(link = "logit")
)

summary(m_bb)

# DHARMa residual diagnostics (titled)
if (have_pkg("DHARMa")) {
  library(DHARMa)
  res_bb <- simulateResiduals(m_bb, plot = FALSE)
  plot(res_bb)
  print(testDispersion(res_bb))
  print(testZeroInflation(res_bb))
}
```

### 3.3 Cluster-robust inference (beta-binomial)

```{r betabin-robust}

# Cluster-robust HC0 (sandwich) from glmmTMB
V_hc0 <- glmmTMB::vcovHC(m_bb, cluster = dat2$co2LF)  # HC0 only

# Stata-style small-sample multiplier you used in Tobit
G <- dplyr::n_distinct(dat2$co2LF)
V_stata <- (G/(G - 1)) * V_hc0

# Fixed effects (conditional) inference using z
b  <- glmmTMB::fixef(m_bb)$cond
Vb <- V_stata[names(b), names(b), drop = FALSE]

se <- sqrt(diag(Vb))
z  <- b / se
p  <- 2 * pnorm(-abs(z))

bb_inf <- data.frame(term = names(b), estimate = b, se_cr = se, z = z, p = p)
bb_inf

```

## 4) Common-scale predictive checks (Tobit vs Beta-binomial)

### 4.1 Predictions: mean share and P(zero)

```{r predictions}
# --- Tobit predictions on original SHARE scale via simulation ---
set.seed(1)
S <- 5000
sd_i <- sigma_hat / sqrt(a)
ystar <- matrix(rnorm(N * S), nrow = N, ncol = S)
ystar <- xb + sd_i * ystar

share_sim <- ifelse(ystar <= cpoint, 0, exp(ystar))
share_hat_tobit <- rowMeans(share_sim)
p0_hat_tobit <- rowMeans(ystar <= cpoint)

# --- Beta-binomial predictions ---
# Mean share (conditional mean) from glmmTMB
p_hat_bb <- as.numeric(predict(m_bb, type = "response"))

# P(k=0) via simulation
set.seed(1)
sim_obj <- simulate(m_bb, nsim = 500)
# simulate(m_bb) returns a data.frame whose columns contain N x 2 matrices [k, n-k]
sim_k_mat <- do.call(cbind, lapply(sim_obj, function(m) m[, 1]))
stopifnot(nrow(sim_k_mat) == N, ncol(sim_k_mat) == 500)
p0_hat_bb <- rowMeans(sim_k_mat == 0)

# Attach predictions to dat2 for safe grouped summaries
# (This prevents the classic weighted.mean length mismatch.)
dat_pred <- dat2 %>%
  mutate(
    tobit_share_hat = share_hat_tobit,
    tobit_p0_hat = p0_hat_tobit,
    bb_share_hat = p_hat_bb,
    bb_p0_hat = p0_hat_bb
  )
```

### 4.2 Overall predictive check table

```{r overall-checks}
obs_share <- dat_pred$p
obs_zero  <- (dat_pred$k == 0)

summ_overall <- data.frame(
  metric = c("mean_share_unw", "mean_share_w_n",
             "zero_rate_unw", "zero_rate_w_n"),
  observed = c(
    mean(obs_share),
    weighted.mean(obs_share, w = dat_pred$n),
    mean(obs_zero),
    weighted.mean(obs_zero, w = dat_pred$n)
  ),
  tobit = c(
    mean(dat_pred$tobit_share_hat),
    weighted.mean(dat_pred$tobit_share_hat, w = dat_pred$n),
    mean(dat_pred$tobit_p0_hat),
    weighted.mean(dat_pred$tobit_p0_hat, w = dat_pred$n)
  ),
  betabinomial = c(
    mean(dat_pred$bb_share_hat),
    weighted.mean(dat_pred$bb_share_hat, w = dat_pred$n),
    mean(dat_pred$bb_p0_hat),
    weighted.mean(dat_pred$bb_p0_hat, w = dat_pred$n)
  )
)

round_df(as.data.frame(summ_overall), 3)
```

### 4.3 Predictive checks by period

```{r by-period}
# IMPORTANT: we summarise on dat_pred so x and w are always the same length in each group.

by_period <- dat_pred %>%
  group_by(period) %>%
  summarise(
    n_obs = n(),
    obs_share_w_n = weighted.mean(p, w = n),
    obs_zero_w_n  = weighted.mean(k == 0, w = n),
    tobit_share_w_n = weighted.mean(tobit_share_hat, w = n),
    tobit_p0_w_n    = weighted.mean(tobit_p0_hat, w = n),
    bb_share_w_n    = weighted.mean(bb_share_hat, w = n),
    bb_p0_w_n       = weighted.mean(bb_p0_hat, w = n)
  )

round_df(as.data.frame(by_period), 3)
```

### 4.4 Calibration plots by period (clearly titled)

```{r calib-plots}
op <- par(no.readonly = TRUE)
par(mfrow = c(2, 2))

# 1) Beta-binomial: zero rate
plot(
  by_period$obs_zero_w_n, by_period$bb_p0_w_n,
  xlab = "Observed zero rate (weighted by n)",
  ylab = "Predicted P(zero) (weighted by n)",
  main = "Beta-binomial (glmmTMB): Zero-rate calibration by period",
  pch = 19
)
add_goodfit_subtitle(good_line_txt)
abline(0, 1, lty = 2)
text(by_period$obs_zero_w_n, by_period$bb_p0_w_n, labels = by_period$period, pos = 4, cex = 0.8)

# 2) Tobit: zero rate
plot(
  by_period$obs_zero_w_n, by_period$tobit_p0_w_n,
  xlab = "Observed zero rate (weighted by n)",
  ylab = "Predicted P(zero) (weighted by n)",
  main = sprintf("Tobit/intreg (cpoint=%s): Zero-rate calibration by period", cpoint),
  pch = 19
)
add_goodfit_subtitle(good_line_txt)
abline(0, 1, lty = 2)
text(by_period$obs_zero_w_n, by_period$tobit_p0_w_n, labels = by_period$period, pos = 4, cex = 0.8)

# 3) Beta-binomial: mean share
plot(
  by_period$obs_share_w_n, by_period$bb_share_w_n,
  xlab = "Observed mean trauma share (weighted by n)",
  ylab = "Predicted mean trauma share (weighted by n)",
  main = "Beta-binomial (glmmTMB): Mean-share calibration by period",
  pch = 19
)
add_goodfit_subtitle("Good fit: points close to 45° line; errors not concentrated in specific periods.")
abline(0, 1, lty = 2)
text(by_period$obs_share_w_n, by_period$bb_share_w_n, labels = by_period$period, pos = 4, cex = 0.8)

# 4) Tobit: mean share
plot(
  by_period$obs_share_w_n, by_period$tobit_share_w_n,
  xlab = "Observed mean trauma share (weighted by n)",
  ylab = "Predicted mean trauma share (weighted by n)",
  main = sprintf("Tobit/intreg (cpoint=%s): Mean-share calibration by period", cpoint),
  pch = 19
)
add_goodfit_subtitle("Good fit: points close to 45° line; no huge outliers (massive over-prediction).")
abline(0, 1, lty = 2)
text(by_period$obs_share_w_n, by_period$tobit_share_w_n, labels = by_period$period, pos = 4, cex = 0.8)

par(op)
```

### 4.5 Zero calibration by deciles

```{r zero-calibration}
make_calibration <- function(p0_hat, obs_zero, w, name) {
  cuts <- unique(quantile(p0_hat, probs = seq(0, 1, 0.1), na.rm = TRUE))
  bin <- cut(p0_hat, breaks = cuts, include.lowest = TRUE)
  tibble(bin = bin, p0_hat = p0_hat, obs_zero = obs_zero, w = w) %>%
    group_by(bin) %>%
    summarise(
      n = n(),
      pred_w = weighted.mean(p0_hat, w, na.rm = TRUE),
      obs_w  = weighted.mean(obs_zero, w, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(model = name)
}

cal_all <- bind_rows(
  make_calibration(dat_pred$tobit_p0_hat, (dat_pred$k == 0), dat_pred$n, "tobit"),
  make_calibration(dat_pred$bb_p0_hat,    (dat_pred$k == 0), dat_pred$n, "betabinomial")
)

round_df(as.data.frame(cal_all), 3)
```

## 5) Tobit vs Beta-binomial: direction + significance table (rounded)

```{r sign-compare}
# Tobit: already has cluster-robust p-values
# Beta-binomial: uses cluster-robust p-values if clubSandwich works, else model-based.

# Prepare Tobit table
term_t <- out_tobit$term
names(term_t) <- NULL

tobit_tbl <- out_tobit %>%
  transmute(
    term,
    est_tobit = estimate,
    p_tobit = p,
    sign_tobit = ifelse(estimate > 0, "+", "-")
  )

# Prepare beta-binomial table (cluster-bootstrap inference)
bb_tbl <- bb_inf %>%
  transmute(
    term,
    est_bb  = estimate,
    p_bb    = p,
    sign_bb = ifelse(estimate > 0, "+", "-")
  )


comp <- inner_join(tobit_tbl, bb_tbl, by = "term") %>%
  mutate(
    agree_sign = (sign_tobit == sign_bb),
    sig_tobit_05 = (p_tobit < 0.05),
    sig_bb_05 = (p_bb < 0.05)
  ) 

round_df(as.data.frame(comp), 3)
```

